Climate-driven phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-06-04 agujournaltemplate.tex
Discussion 95 % done

% Figure 4
Taking the microscopy community data interpolated within the top 100 meters, we aimed to identify any apparent structure within the time series that could be related to the physical environment. To achieve this, we employed clustering analysis on both the annual aggregate of cell counts from microscopy and the individual monthly measurements. Figure \ref{fig:clustering} $a$ presents the results of clustering the annual aggregate communities using a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster encompassing the years 1996-2004 and 2014-2016, while the second cluster includes the intermediate years from 2004-2013. 
Inter-annual species turnover is illustrated in Figure \ref{fig:clustering} $b$, revealing an increasing trend in turnover until 2004. Turnover subsequently decreased during the period of cluster 2 but increased again in 2014, marking the return to cluster 1. This pattern is also reflected in the hierarchical clustering results in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster respectively. The diatom Thalassiosira dominated the early period and was abundant in all other periods. Pseudo-nitzschia dominates cluster 2, where the haptophyte Emiliana was the second most abundant genus. The later part of cluster 1 is dominated by Emiliana.

@@

% Clearly and briefly restating main findings 
% Interpret what these findings mean in the context of the study
% Emphasize what is novel and important (without comparison)
% Only in following paragraphs: Compare to existing literature, highlight agreements, contradictions, and new persepctives. Discuss why and explain reasons for either.

@@

\subsection{Drivers of change: Gradient forest analysis}
% P4: Introduce GF Analysis and discuss results
To identify the variables that can best predict changes in the phytoplankton community, we conducted a gradient forest analysis using microscopy cell counts at the genus level as predicted variables and climate indices, along with in situ measurements of bottom-up drivers, as predictors. Gradient forest is an extension of the random forest algorithm, designed to identify thresholds and patterns in species abundance or composition along environmental gradients \cite{ellis_gradient_2012}. Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series. Interestingly, our analysis revealed that the strongest predictor of changes within the community was not an in situ measurement but rather the AMO climate index. The AMO is followed in importance by SST, the MEI v.2 index, and nitrate concentration (see Figure \ref{fig:GF}).

These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separates the two yearly aggregate community clusters is NMDS 1, which co-varies most strongly with changes in the AMO, MEI v.2, salinity, evaporation, and precipitation (see Figure \ref{fig:clustering} $c$). Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS 2. This indicates that the climate indices capture some variability that is not represented in the more direct measurements of upwelling intensity. Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, show greater variance along NMDS 1. Salinity, evaporation, and silicate concentration exhibit some explanatory power in the gradient forest analysis, although they are weaker predictors than SST, nitrate concentration, and the climate indices.

Using microscopy cell counts as a basis for this analysis means that the results should be interpreted in terms of changes in the microphytoplankton community, due to the size limitations of identifying species via light microscopy. Grouping cell counts by genera does create a loss of resolution but is based on the assumption that species within a genus generally exhibit similar eco-physiological responses. The output of the gradient forest model is further grouped by broad functional types to facilitate the identification of shifts and trends. The phytoplankton genera that can be best predicted are mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (see Figure \ref{fig:GF} $b$). Within these groups, we observe differing threshold responses in cumulative importance along the predictor ranges.

% P5: Climate impacts
Although regional effects may vary, a positive phase in the AMO is generally associated with periods of warmer temperatures and a shift of the ITCZ from north to south. This shift of the ITCZ southward increases wind-driven upwelling in the Cariaco Basin \cite{taylor_ecosystem_2012}. The threshold responses between functional groups, which relate to shifts in cell count abundances, show contrasting patterns. Diatoms and cyanobacteria exhibit strong response thresholds for negative AMO anomalies. This interesting threshold response observed for the AMO, which, if sufficiently negative, seems to correlate with a reorganization of the physical regime in the Cariaco Basin. The abundance of haptophytes shows a strong threshold at positive AMO anomalies, which could be related to the pattern observed from HPLC data that haptophytes predominantly exhibit blooms above the mixed layer and at a lower frequency than other functional groups \cite{pinckney_phytoplankton_2015}.

% Move the general point on ENSO to intro! Only Discussion-relevant part here..
The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea. During ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin, and only \qtyrange{24}{36}{\%} explanation of variance with a time lag of 12 months using linear regression. In contrast, we found relatively strong predictive power for changes in the phytoplankton community with a 4-month time lag. We also tested a 12-month lag and found no significant improvement.
The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for a negative MEI v.2 index, which suggests La Niña conditions. For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events. Diatoms do not exhibit a clear threshold response for the MEI v.2 index but are affected across the entire range. Interestingly, positive anomalies of the MEI v.2, representing strong ENSO events, are found mostly in Cluster 1 (see Figure \ref{fig:clustering} $e$).

The in-situ measurements that best predict changes in the microphytoplankton community were SST, nitrate concentration, and salinity, in that order. Nitrate is a crucial nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}. For dinoflagellates and diatoms, we observe a strong threshold response around \qty{7}{\micro \mole}, which indicates that strong upwelling conditions drive blooms of these functional groups. Other functional groups, including \textit{Mesodinium} and \textit{Eutreptria} also show a threshold response at even higher nitrate concentrations (around \qty{9}{\micro \mole}), indicating an effect on their abundance from very strong upwelling. Similarly, for temperature, dinoflagellates and diatoms exhibit a threshold around \qtyrange{21}{23}{\celsius}, which indicates higher abundances during low temperatures in the upper \qty{100}{m} driven by strong upwelling. For the other predictors, we do not observe strong threshold responses.

Gradient forest analysis clearly identified the AMO and MEI v.2 climate indices as the best predictors of changes in the microphytoplankton community in the Cariaco Basin. This is an interesting result, as one would expect that more direct measures of ecosystem variability could better predict shifts in abundances. One possible explanation is that in situ measurements contain significantly stronger variability due to measurement error, as well as local and temporal variability, and generally provide only a single snapshot for an entire month. Climate indices are detrended and integrated measurements of long-term changes that are matching in scale to the dynamics of long-term changes within a phytoplankton community. However, we do not see such strong predictive power in the ERA5 climate reanalysis-derived mean wind speed, which also aggregates measurements and model output over an entire month. 
Another explanation could be that the climate indices capture a dimension of variability not found in our in-situ measurements, which is also represented in our NMDS clustering (see Figure \ref{fig:clustering} $c$), where both MEI v.2 and AMO scale along the second axis (NMDS 2). %TODO test more of my variables along these axes, then add to discussion here and maybe put plot in supplemental material.

@@

% FROM FRAKER ET AL. 2022
%"Note that although this analysis identifies predictive relationships between drivers and the biological indicators, these relationships do not necessarily imply mechanistic relationships. As such, this analysis represents the first stages of an overall analysis (a narrowing of the potential key relationships and when they matter from a much larger dataset), and additional statistical analyses, modeling, and experimental work may be required to develop a full mechanistic understanding of driver-response relationships and ecosystem change."

- alongside a global reduction in frequency compared to the linear trend between 2006 and 2012 \cite{dai_coastal_2023}. % <- for this paper, check references to ENSO MEI v.2!!! This could be interesting!

Our limitation to bottom-up drivers of community change could also limit the explanatory power of our analysis, as top-down processes such as zooplankton grazing strongly affect the structure of the phytoplankton community \cite{banas_adding_2011, acevedo-trejos_mechanisms_2015}. In the Cariaco basin in particular, a trophic cascade from the collapse of sardine fisheries through an increasing abundance of mesozooplaknton has been proposed to have affected the phytoplankton community \cite{muller-karger_scientific_2019}. Due to the lack of temporal resolution for higher trophic levels and the lack of zooplankton data for the first 6 years of the time series, we propose that mechanistic modeling studies could be well suited to investigate this aspect of the CARIACO time series further.

@@

% Highlight that the climate indices explain changes in the community and abundance, but not necessarily changed in diversity (as there is no sig diff beteween clusters). Probably need more data, other approaches to untangle effects on diversity.
In this study, we have found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco basin. Despite multiannual trends in warming, reduced productivity and a collapse in fisheries and phytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions returned. We see these patterns clearly in the community data retrieved from microscopy, which shows the value of the arduous work of identifying and counting phytoplankton cells. The strongest bottom-up variable predicting changes in the phytoplankton community is the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local ecosystems through low-frequency natural cycles. We hope that the work hints at the resilient nature of phytoplankton communities under global change, which should be further investigated by studying long-term ocean time series.

>>>>

% Figure 4
Taking the microscopy community data interpolated within the top 100 meters, we aimed to identify any apparent structure within the time series that could be related to the physical environment. To achieve this, we employed clustering analysis on both the annual aggregate of cell counts from microscopy and the individual monthly measurements. Figure \ref{fig:clustering} $a$ presents the results of clustering the annual aggregate communities using a linear complete linkage algorithm based on the binary Jaccard matrix. The community is divided into two groups, with the first cluster encompassing the years 1996-2004 and 2014-2016, while the second cluster includes the intermediate years from 2004-2013. 
Inter-annual species turnover is illustrated in Figure \ref{fig:clustering} $b$, revealing an increasing trend in turnover until 2004. Turnover subsequently decreased during the period of cluster 2 but increased again in 2014, marking the return to cluster 1. This pattern is also reflected in the hierarchical clustering results in Figure \ref{fig:clustering} $a$, where cluster 2 shows a lower distance in the clustering tree between the years. We additionally extracted the three dominant genera for the second cluster and the early and late parts of the first cluster respectively. The diatom \textit{Thalassiosira} dominated the early period and was abundant in all other periods. \textit{Pseudo-nitzschia} dominates cluster 2, where the haptophyte \textit{Emiliana} was the second most abundant genus. The later part of cluster 1 is dominated by \textit{Emiliana}.

@@

% Subhendu's Comment:
% Clearly and briefly restating main findings 
% Interpret what these findings mean in the context of the study
% Emphasize what is novel and important (without comparison)
% Only in following paragraphs: Compare to existing literature, highlight agreements, contradictions, and new persepctives. Discuss why and explain reasons for either.

@@

\subsection{Drivers of change: Gradient forest analysis}
% P4: Introduce GF Analysis and discuss results
To identify the variables that best predict changes in the phytoplankton community, we conducted a gradient forest analysis using time series of climate indices and in situ measurements to predict genus-level abundances from microscopy cell counts. Gradient forest is an extension of the random forest algorithm, designed to identify thresholds and patterns in community composition along environmental gradients \cite{ellis_gradient_2012}. Limiting the analysis to bottom-up drivers ensured the greatest possible data coverage throughout the CARIACO time series. Interestingly, our analysis revealed that the strongest predictor of changes within the community was not an in situ measurement but rather the AMO climate index. The AMO is followed in importance by SST, the MEI v.2 index, and nitrate concentration (see Figure \ref{fig:GF}).

% P5: Climate impacts
Although regional effects may vary, a negative phase in the AMO is generally associated with periods of colder temperatures and a southward shift of the ITCZ \cite{knight_climate_2006, colna2017latitudinal}. In the Cariaco Basin, a southward shift of the ITCZ correlates with increased wind-driven upwelling \cite{taylor_ecosystem_2012}. The threshold responses between functional groups, which relate to shifts in cell count abundances, show contrasting patterns. Diatoms, dinoflagellates, and haptophytes exhibit strong response thresholds for negative AMO anomalies, which could point to the effects of the resulting stronger upwelling conditions. The abundance of haptophytes additionally shows a threshold at positive AMO anomalies, which may be related to the pattern observed from HPLC data that haptophytes predominantly exhibit blooms above the mixed layer and at a lower frequency than other functional groups \cite{pinckney_phytoplankton_2015}.

The El Niño Southern Oscillation (ENSO) has a strong teleconnection to the Caribbean Sea. During ENSO (indicated by a positive MEI v.2 index) sea surface temperatures rise, and trade winds blow more to the north-northwest \cite{enfield_tropical_1997}. \citeA{taylor_ecosystem_2012} found no strong correlation between the MEI v.2 and measurements in the Cariaco Basin and only \qtyrange{24}{36}{\%} explanation of variance using linear regression with a time lag of 12 months. In contrast, we found relatively strong predictive power for changes in the phytoplankton community for the MEI v.2 index with a 4-month time lag. We also tested a 12-month lag and found no significant improvement (see Table \ref{tab:sup:GFlagtests} in the appendix).
The threshold responses indicate a shift in the abundances of dinoflagellates and haptophytes for negative MEI v.2 index anomalies, which represent La Niña conditions. For cyanobacteria and haptophytes, we observed a pronounced threshold response to positive anomalies, demonstrating sensitivity to ENSO events. Diatoms do not exhibit a clear threshold response for the MEI v.2 index but are affected across the entire range. Interestingly, positive anomalies of the MEI v.2, representing strong ENSO events, are mostly found in Cluster 1 (see Figure \ref{fig:clustering} $e$).

The in-situ measurements that best predict changes in the microphytoplankton community were SST, nitrate concentration, and salinity, in that order. Nitrate is a crucial nutrient for phytoplankton growth and has been found to be the growth-limiting nutrient during stratification periods in the Cariaco Basin \cite{muller-karger_scientific_2019}. For diatoms, we observe a strong threshold response above \qty{7}{\micro \molar}, indicating that intense upwelling conditions drive blooms, resulting in predictable changes in cell counts. Similarly, for SST, haptophytes and diatoms exhibit a threshold around \qtyrange{22}{24}{\celsius}, indicating thresholds at low surface temperatures driven by stronger upwelling.

These results match the observed NMDS clustering of the monthly cell counts in relation to the predictor variables, as the scaling axis that separates the two yearly aggregate community clusters is NMDS 1, which covaries most strongly with changes in the AMO, MEI v.2, salinity, evaporation, and precipitation (see Figure \ref{fig:clustering} $c$). Variables directly related to upwelling conditions, such as temperature, \qty{21}{\celsius} isotherm, wind speed, and nutrient concentrations vary more strongly along NMDS 2. Salinity and silicate concentration, which are also strongly related to water mass exchanges through upwelling but additionally depend on freshwater input from precipitation and surface runoff, show greater variance along NMDS 1. Salinity, evaporation, and silicate concentration exhibit some explanatory power in the gradient forest analysis, although they are weaker predictors than SST, nitrate concentration, and the climate indices (see Figure \ref{fig:GF}). This indicates that the climate indices capture variability not represented in the more direct measurements of upwelling intensity.

@@

% 1.  Discuss Limitations of Light microscopy cell counts in CARIACO Time Series
Using microscopy cell counts as a basis for this analysis means that the results should be interpreted in terms of changes in the microphytoplankton community, due to the size limitations of identifying species via light microscopy. Grouping cell counts by genus does create a loss of resolution but is based on the assumption that species within a genus generally exhibit similar eco-physiological responses. The output of the gradient forest model is further grouped by broad functional types to facilitate the identification of shifts and trends. The phytoplankton genera that can be best predicted are mostly diatoms, along with some haptophytes, dinoflagellates, and cyanobacteria (see Figure \ref{fig:GF} $b$).
This is very likely not a complete picture of the phytoplankton community -> compare to HPLC data.

% 2. Discuss limitations of GF analysis
% FROM FRAKER ET AL. 2022
%"Note that although this analysis identifies predictive relationships between drivers and the biological indicators, these relationships do not necessarily imply mechanistic relationships. As such, this analysis represents the first stages of an overall analysis (a narrowing of the potential key relationships and when they matter from a much larger dataset), and additional statistical analyses, modeling, and experimental work may be required to develop a full mechanistic understanding of driver-response relationships and ecosystem change."
Without adjusting for time lags, SST and nitrate concentration are the strongest predictors of community change (see Figure \ref{fig:sup:GFoutput_nolags} in the appendix). Incorporating time lags slightly improves the overall predictive strength, as quantified by the $R^2$ weighted importance, although both SST and the AMO index as the strongest predictors yield a relatively low value around \qty{0.03}{}. This is to be expected with the data we are working with, and in line with values reported in other studies utilizing the gradient forest algorithm \cite{}.

- There is some correction for correlation between the predictors, and we check the time lags to test for impeding effects of autocorrelation, but still the model does not take the temporal aspect into account in building the random forest models underlying the gradient forest analysis. It can however detect non-linear responses between variables.

% 3. Limitation to bottom-up drivers and climate indices
Our limitation to bottom-up drivers of community change could also limit the explanatory power of our analysis, as top-down processes such as zooplankton grazing strongly affect the structure of the phytoplankton community \cite{banas_adding_2011, acevedo-trejos_mechanisms_2015}. In the Cariaco Basin in particular, a trophic cascade from the collapse of sardine fisheries through an increasing abundance of mesozooplankton has been proposed to have affected the phytoplankton community \cite{muller-karger_scientific_2019}. Due to the lack of temporal resolution for higher trophic levels and the absence of zooplankton data for the first six years of the time series, we propose that mechanistic modeling studies could be well-suited to investigate this aspect of the CARIACO time series further.

@@

In this study, we found a strong impact of decadal-scale climatic oscillations on shifts in the phytoplankton community in the tropical coastal ecosystem of the Cariaco Basin.

Despite multiannual trends in warming, reduced productivity, and a collapse in fisheries and phytoplankton cell counts, the phytoplankton community returned to previous levels once the climatic and local physical conditions were restored.

We observe these patterns clearly in the community data obtained from microscopy, which demonstrates the value of the rigorous work involved in identifying and counting phytoplankton cells.

The strongest bottom-up variable predicting changes in the phytoplankton community is the AMO index, highlighting the complex nature of climate-driven ecosystem variability and the connections between global climate and local ecosystems through low-frequency natural cycles. We can concur with \citeA{nye_ecosystem_2014} that the AMO appears to be a suitable candidate as an indicator of ecosystem state, particularly for the phytoplankton community of the Cariaco Basin.

The analysis could be extended by including top-down effects on the phytoplankton community, particularly zooplankton and fisheries data. 
We hope that the work hints at the resilient nature of phytoplankton communities under global change, which should be further investigated by studying long-term ocean time series.
